---
title: "Clustering / Differential Gene Expression Analysis"
output: pdf_document
author: Lara Feulner
---

# Original analysis performed on M2 Macbook Air (2022) running macOS Monterey 12.4 and R 4.2.3.

# Notable package versions used for analysis: Seurat 4.1.1, SeuratObject 4.1.0, sctransform 0.3.3, uwot 0.1.11, irlba 2.3.5. See full list on this GitHub for more details.  

# Note this has not been tested on Seurat v5 and may need modifications, can also expect changes in differential expression results compared to v4.

```{r}
# Load packages

library(Seurat)
```

```{r}
# Load filtered count matrices for timepoint replicates

E14.r1 <- Read10X("~/FilteredCounts/E14.5_r1_counts_mat")
E14.r2 <- Read10X("~/FilteredCounts/E14.5_r2_counts_mat")
E14.r3 <- Read10X("~/FilteredCounts/E14.5_r3_counts_mat")
E14.r4 <- Read10X("~/FilteredCounts/E14.5_r4_counts_mat")
E14.r5 <- Read10X("~/FilteredCounts/E14.5_r5_counts_mat")
E14.r6 <- Read10X("~/FilteredCounts/E14.5_r6_counts_mat")

E16.r1 <- Read10X("~/FilteredCounts/E16.5_r1_counts_mat")
E16.r2 <- Read10X("~/FilteredCounts/E16.5_r2_counts_mat")
E16.r3 <- Read10X("~/FilteredCounts/E16.5_r3_counts_mat")
E16.r4 <- Read10X("~/FilteredCounts/E16.5_r4_counts_mat")
E16.r5 <- Read10X("~/FilteredCounts/E16.5_r5_counts_mat")
E16.r6 <- Read10X("~/FilteredCounts/E16.5_r6_counts_mat")

E18.r1 <- Read10X("~/FilteredCounts/E18.5_r1_counts_mat")
E18.r2 <- Read10X("~/FilteredCounts/E18.5_r2_counts_mat")
E18.r3 <- Read10X("~/FilteredCounts/E18.5_r3_counts_mat")
E18.r4 <- Read10X("~/FilteredCounts/E18.5_r4_counts_mat")
E18.r5 <- Read10X("~/FilteredCounts/E18.5_r5_counts_mat")
E18.r6 <- Read10X("~/FilteredCounts/E18.5_r6_counts_mat")

P0.r1 <- Read10X("~/FilteredCounts/P0_r1_counts_mat")
P0.r2 <- Read10X("~/FilteredCounts/P0_r2_counts_mat")
P0.r3 <- Read10X("~/FilteredCounts/P0_r3_counts_mat")
P0.r4 <- Read10X("~/FilteredCounts/P0_r4_counts_mat")
P0.r5 <- Read10X("~/FilteredCounts/P0_r5_counts_mat")
P0.r6 <- Read10X("~/FilteredCounts/P0_r6_counts_mat")

P4.r1 <- Read10X("~/FilteredCounts/P4_r1_counts_mat")
P4.r2 <- Read10X("~/FilteredCounts/P4_r2_counts_mat")
P4.r3 <- Read10X("~/FilteredCounts/P4_r3_counts_mat")
P4.r4 <- Read10X("~/FilteredCounts/P4_r4_counts_mat")
P4.r5 <- Read10X("~/FilteredCounts/P4_r5_counts_mat")
P4.r6 <- Read10X("~/FilteredCounts/P4_r6_counts_mat")

P7.r1 <- Read10X("~/FilteredCounts/P7_r1_counts_mat")
P7.r2 <- Read10X("~/FilteredCounts/P7_r2_counts_mat")
P7.r3 <- Read10X("~/FilteredCounts/P7_r3_counts_mat")
P7.r4 <- Read10X("~/FilteredCounts/P7_r4_counts_mat")
P7.r5 <- Read10X("~/FilteredCounts/P7_r5_counts_mat")
P7.r6 <- Read10X("~/FilteredCounts/P7_r6_counts_mat")
```

```{r}
# Create Seurat objects

E14.r1 <- CreateSeuratObject(E14.r1, names.field=c(1,2))
E14.r2 <- CreateSeuratObject(E14.r2, names.field=c(1,2))
E14.r3 <- CreateSeuratObject(E14.r3, names.field=c(1,2))
E14.r4 <- CreateSeuratObject(E14.r4, names.field=c(1,2))
E14.r5 <- CreateSeuratObject(E14.r5, names.field=c(1,2))
E14.r6 <- CreateSeuratObject(E14.r6, names.field=c(1,2))

E16.r1 <- CreateSeuratObject(E16.r1, names.field=c(1,2))
E16.r2 <- CreateSeuratObject(E16.r2, names.field=c(1,2))
E16.r3 <- CreateSeuratObject(E16.r3, names.field=c(1,2))
E16.r4 <- CreateSeuratObject(E16.r4, names.field=c(1,2))
E16.r5 <- CreateSeuratObject(E16.r5, names.field=c(1,2))
E16.r6 <- CreateSeuratObject(E16.r6, names.field=c(1,2))

E18.r1 <- CreateSeuratObject(E18.r1, names.field=c(1,2))
E18.r2 <- CreateSeuratObject(E18.r2, names.field=c(1,2))
E18.r3 <- CreateSeuratObject(E18.r3, names.field=c(1,2))
E18.r4 <- CreateSeuratObject(E18.r4, names.field=c(1,2))
E18.r5 <- CreateSeuratObject(E18.r5, names.field=c(1,2))
E18.r6 <- CreateSeuratObject(E18.r6, names.field=c(1,2))

P0.r1 <- CreateSeuratObject(P0.r1, names.field=c(1,2))
P0.r2 <- CreateSeuratObject(P0.r2, names.field=c(1,2))
P0.r3 <- CreateSeuratObject(P0.r3, names.field=c(1,2))
P0.r4 <- CreateSeuratObject(P0.r4, names.field=c(1,2))
P0.r5 <- CreateSeuratObject(P0.r5, names.field=c(1,2))
P0.r6 <- CreateSeuratObject(P0.r6, names.field=c(1,2))

P4.r1 <- CreateSeuratObject(P4.r1, names.field=c(1,2))
P4.r2 <- CreateSeuratObject(P4.r2, names.field=c(1,2))
P4.r3 <- CreateSeuratObject(P4.r3, names.field=c(1,2))
P4.r4 <- CreateSeuratObject(P4.r4, names.field=c(1,2))
P4.r5 <- CreateSeuratObject(P4.r5, names.field=c(1,2))
P4.r6 <- CreateSeuratObject(P4.r6, names.field=c(1,2))

P7.r1 <- CreateSeuratObject(P7.r1, names.field=c(1,2))
P7.r2 <- CreateSeuratObject(P7.r2, names.field=c(1,2))
P7.r3 <- CreateSeuratObject(P7.r3, names.field=c(1,2))
P7.r4 <- CreateSeuratObject(P7.r4, names.field=c(1,2))
P7.r5 <- CreateSeuratObject(P7.r5, names.field=c(1,2))
P7.r6 <- CreateSeuratObject(P7.r6, names.field=c(1,2))
```

```{r}
# Calculate percent mitochondria

E14.r1[["percent_mito"]] <- PercentageFeatureSet(E14.r1, pattern = "^mt-")
E14.r2[["percent_mito"]] <- PercentageFeatureSet(E14.r2, pattern = "^mt-")
E14.r3[["percent_mito"]] <- PercentageFeatureSet(E14.r3, pattern = "^mt-")
E14.r4[["percent_mito"]] <- PercentageFeatureSet(E14.r4, pattern = "^mt-")
E14.r5[["percent_mito"]] <- PercentageFeatureSet(E14.r5, pattern = "^mt-")
E14.r6[["percent_mito"]] <- PercentageFeatureSet(E14.r6, pattern = "^mt-")

E16.r1[["percent_mito"]] <- PercentageFeatureSet(E16.r1, pattern = "^mt-")
E16.r2[["percent_mito"]] <- PercentageFeatureSet(E16.r2, pattern = "^mt-")
E16.r3[["percent_mito"]] <- PercentageFeatureSet(E16.r3, pattern = "^mt-")
E16.r4[["percent_mito"]] <- PercentageFeatureSet(E16.r4, pattern = "^mt-")
E16.r5[["percent_mito"]] <- PercentageFeatureSet(E16.r5, pattern = "^mt-")
E16.r6[["percent_mito"]] <- PercentageFeatureSet(E16.r6, pattern = "^mt-")

E18.r1[["percent_mito"]] <- PercentageFeatureSet(E18.r1, pattern = "^mt-")
E18.r2[["percent_mito"]] <- PercentageFeatureSet(E18.r2, pattern = "^mt-")
E18.r3[["percent_mito"]] <- PercentageFeatureSet(E18.r3, pattern = "^mt-")
E18.r4[["percent_mito"]] <- PercentageFeatureSet(E18.r4, pattern = "^mt-")
E18.r5[["percent_mito"]] <- PercentageFeatureSet(E18.r5, pattern = "^mt-")
E18.r6[["percent_mito"]] <- PercentageFeatureSet(E18.r6, pattern = "^mt-")

P0.r1[["percent_mito"]] <- PercentageFeatureSet(P0.r1, pattern = "^mt-")
P0.r2[["percent_mito"]] <- PercentageFeatureSet(P0.r2, pattern = "^mt-")
P0.r3[["percent_mito"]] <- PercentageFeatureSet(P0.r3, pattern = "^mt-")
P0.r4[["percent_mito"]] <- PercentageFeatureSet(P0.r4, pattern = "^mt-")
P0.r5[["percent_mito"]] <- PercentageFeatureSet(P0.r5, pattern = "^mt-")
P0.r6[["percent_mito"]] <- PercentageFeatureSet(P0.r6, pattern = "^mt-")

P4.r1[["percent_mito"]] <- PercentageFeatureSet(P4.r1, pattern = "^mt-")
P4.r2[["percent_mito"]] <- PercentageFeatureSet(P4.r2, pattern = "^mt-")
P4.r3[["percent_mito"]] <- PercentageFeatureSet(P4.r3, pattern = "^mt-")
P4.r4[["percent_mito"]] <- PercentageFeatureSet(P4.r4, pattern = "^mt-")
P4.r5[["percent_mito"]] <- PercentageFeatureSet(P4.r5, pattern = "^mt-")
P4.r6[["percent_mito"]] <- PercentageFeatureSet(P4.r6, pattern = "^mt-")

P7.r1[["percent_mito"]] <- PercentageFeatureSet(P7.r1, pattern = "^mt-")
P7.r2[["percent_mito"]] <- PercentageFeatureSet(P7.r2, pattern = "^mt-")
P7.r3[["percent_mito"]] <- PercentageFeatureSet(P7.r3, pattern = "^mt-")
P7.r4[["percent_mito"]] <- PercentageFeatureSet(P7.r4, pattern = "^mt-")
P7.r5[["percent_mito"]] <- PercentageFeatureSet(P7.r5, pattern = "^mt-")
P7.r6[["percent_mito"]] <- PercentageFeatureSet(P7.r6, pattern = "^mt-")
```

```{r}
# Add timepoint metadata

E14.r1$Age <- "E14.5"
E14.r2$Age <- "E14.5"
E14.r3$Age <- "E14.5"
E14.r4$Age <- "E14.5"
E14.r5$Age <- "E14.5"
E14.r6$Age <- "E14.5"

E16.r1$Age <- "E16.5"
E16.r2$Age <- "E16.5"
E16.r3$Age <- "E16.5"
E16.r4$Age <- "E16.5"
E16.r5$Age <- "E16.5"
E16.r6$Age <- "E16.5"

E18.r1$Age <- "E18.5"
E18.r2$Age <- "E18.5"
E18.r3$Age <- "E18.5"
E18.r4$Age <- "E18.5"
E18.r5$Age <- "E18.5"
E18.r6$Age <- "E18.5"

P0.r1$Age <- "P0"
P0.r2$Age <- "P0"
P0.r3$Age <- "P0"
P0.r4$Age <- "P0"
P0.r5$Age <- "P0"
P0.r6$Age <- "P0"

P4.r1$Age <- "P4"
P4.r2$Age <- "P4"
P4.r3$Age <- "P4"
P4.r4$Age <- "P4"
P4.r5$Age <- "P4"
P4.r6$Age <- "P4"

P7.r1$Age <- "P7"
P7.r2$Age <- "P7"
P7.r3$Age <- "P7"
P7.r4$Age <- "P7"
P7.r5$Age <- "P7"
P7.r6$Age <- "P7"
```

```{r}
# Add sex metadata where M=male and F=female

E14.r1$Sex <- "M"
E14.r2$Sex <- "M"
E14.r3$Sex <- "F"
E14.r4$Sex <- "M"
E14.r5$Sex <- "F"
E14.r6$Sex <- "F"

E16.r1$Sex <- "M"
E16.r2$Sex <- "F"
E16.r3$Sex <- "F"
E16.r4$Sex <- "M"
E16.r5$Sex <- "F"
E16.r6$Sex <- "F"

E18.r1$Sex <- "M"
E18.r2$Sex <- "M"
E18.r3$Sex <- "F"
E18.r4$Sex <- "M"
E18.r5$Sex <- "F"
E18.r6$Sex <- "M"

P0.r1$Sex <- "F"
P0.r2$Sex <- "F"
P0.r3$Sex <- "M"
P0.r4$Sex <- "F"
P0.r5$Sex <- "M"
P0.r6$Sex <- "F"

P4.r1$Sex <- "F"
P4.r2$Sex <- "M"
P4.r3$Sex <- "M"
P4.r4$Sex <- "F"
P4.r5$Sex <- "M"
P4.r6$Sex <- "F"

P7.r1$Sex <- "M"
P7.r2$Sex <- "F"
P7.r3$Sex <- "M"
P7.r4$Sex <- "F"
P7.r5$Sex <- "F"
P7.r6$Sex <- "M"
```

```{r}
# Merge all timepoint replicates

data = merge(E14.r1, y=c(E14.r2, E14.r3, E14.r4, E14.r5, E14.r6, E16.r1, E16.r2, E16.r3, E16.r4, E16.r5, E16.r6, E18.r1, E18.r2, E18.r3, E18.r4, E18.r5, E18.r6, P0.r1, P0.r2, P0.r3, P0.r4, P0.r5, P0.r6, P4.r1, P4.r2, P4.r3, P4.r4, P4.r5, P4.r6, P7.r1, P7.r2, P7.r3, P7.r4, P7.r5, P7.r6))
```

```{r}
# Remove replicate R objects to clear space

remove(E14.r1)
remove(E14.r2)
remove(E14.r3)
remove(E14.r4)
remove(E14.r5)
remove(E14.r6)

remove(E16.r1)
remove(E16.r2)
remove(E16.r3)
remove(E16.r4)
remove(E16.r5)
remove(E16.r6)

remove(E18.r1)
remove(E18.r2)
remove(E18.r3)
remove(E18.r4)
remove(E18.r5)
remove(E18.r6)

remove(P0.r1)
remove(P0.r2)
remove(P0.r3)
remove(P0.r4)
remove(P0.r5)
remove(P0.r6)

remove(P4.r1)
remove(P4.r2)
remove(P4.r3)
remove(P4.r4)
remove(P4.r5)
remove(P4.r6)

remove(P7.r1)
remove(P7.r2)
remove(P7.r3)
remove(P7.r4)
remove(P7.r5)
remove(P7.r6)
```

```{r}
# Load cell cycle genes 

s_mouse_genes_regev <- read.table("~/mouse_s_genes.csv", sep="\t", header=T)
s_mouse_genes_regev <- s_mouse_genes_regev$s_mouse

g2m_mouse_genes_regev <- read.table("~/mouse_g2m_genes.csv", sep="\t", header=T)
g2m_mouse_genes_regev <- g2m_mouse_genes_regev$g2m_mouse
```

```{r}
# Perform normalization and cell cycle scoring, stored in a new Seurat object (data2)

data2 <- NormalizeData(data, normalization.method = 'LogNormalize', scale.factor = 1e4)
data2 <- CellCycleScoring(data2, s.features = s_mouse_genes_regev, g2m.features = g2m_mouse_genes_regev)
```

```{r}
# Transfer cell cycle scoring information from the second Seurat object to the original object
# As the original object has not been normalized, SCTransform can be subsequently used

data$S.Score <- data2@meta.data$S.Score
data$G2M.Score <- data2@meta.data$G2M.Score
data$Phase <- data2@meta.data$Phase
```

```{r}
# Perform SCTransform

options(future.globals.maxSize=4000000000)
# Otherwise may get error similar to below when running SCTransform: 
# The total size of the 19 globals exported for future expression (‘FUN()’) is 1.35 GiB. This exceeds the maximum allowed size 500.00 MiB per plan() argument 'maxSizeOfObjects'. 
data_SCT <- SCTransform(data, vars.to.regress = c("S.Score", "G2M.Score", "percent_mito"))
```

# Below are the codes used for subsequent dimension reduction and clustering, but note there are minor variances in PCA results across different environments (ex. different machines) that result in similar but not identical UMAPs even using same package versions. Issue discussed here: https://github.com/satijalab/seurat/issues/5358 and https://github.com/satijalab/seurat/issues/8501

```{r}
# Run PCA

data_PCA <- RunPCA(object = data_SCT, npcs = 100)
```

```{r}
# Run UMAP and Find Neighbors
# Note that very large dims were found required to separate pericytes from both smooth muscle clusters 

data_UMAP <- RunUMAP(object = data_PCA, reduction = "pca", dims = c(1:83))
data_neighbors <- FindNeighbors(data_UMAP, compute.SNN=T, reduction="pca", c(1:83))
```

```{r}
# Perform Clustering
# Note this was repeated at higher resolutions (up to resolution=10) to help identify smaller cell types and subtypes

data_cluster <- FindClusters(
  object = data_neighbors,
  resolution = 0.3,
  algorithm = 2,
  n.iter = 50,
  n.start = 20
)
```

# To get original UMAP, can transfer dimension reduction information from original Seurat object to data_SCT object. Note that data_SCT should have been generated with sctransform version 0.33 to ensure reproducibility.

```{r}
data_orig_anal <- readRDS("~/Perinatal_clusterannot.Rds")

# If you want to verify matrices following SCTransform are identical:
# sct_data1 <- data_orig_anal@assays$SCT@data
# sct_data2 <- data_cluster@assays$SCT@data
# all.equal(sct_data1, sct_data2)
# if not TRUE need to use data_orig_anal object instead to replicate all downstream plots/analyses!

# Transfer dimension reduction information
data_cluster@reductions$pca = data_orig_anal@reductions$pca
data_cluster@reductions$umap = data_orig_anal@reductions$umap
```

# Now also transfer metadata from original Seurat object to obtain cell type annotation and original clustering information

```{r}
data_cluster@meta.data = data_orig_anal@meta.data
Idents(data_cluster) <- data_cluster@meta.data$Celltype_abbr
```

# Visualization

```{r}
# View Dim plot (see Figure1_plots.Rmd for reproduction)
DimPlot(data_cluster, label=T)
```

```{r}
# View Feature plot (see Figure1_plots.Rmd for reproduction)
FeaturePlot(data_cluster, "Myl2")
```


# Differential Gene Expression Analysis

```{r}
# By cluster (example cluster 13, repeat for all)
# These results were used to guide manual cluster annotation

DefaultAssay(data_cluster) = "SCT"
Idents(data_cluster) <- data_cluster@meta.data$seurat_clusters

cluster13markers <- FindMarkers(data_cluster, ident.1=13, min.pct=0.1, min.diff.pct=0.1, only.pos=T)
```

```{r}
# By cell type (example EPI, repeat for all types)
# Corresponds to Table 2

DefaultAssay(data_cluster) = "SCT"
Idents(data_cluster) <- data_cluster@meta.data$Celltype_abbr

EPImarkers <- FindMarkers(data_cluster, ident.1="EPI", min.pct=0.1, min.diff.pct=0.1, only.pos=T)

# Genes enriched in ACM1 compared to ACM2
# Corresponds to Table 3
# Lower thresholds needed due to low Pitx2 expression at timepoints analysed

ACM1vACM2markers <- FindMarkers(data_cluster, ident.1="ACM1", ident.2="ACM2", min.pct=0.1, logfc.threshold = 0.1, only.pos=F)
```

